Space astrometry of the very massive $\sim$$150\,M_\odot$ candidate runaway star VFTS682

### 2018-08-28 MNRASformat/VFTS682_MNRASL.tex
proof-read

How very massive stars form is still an open question in
astrophysics.  VFTS682 is among the most massive stars known,
with an inferred initial mass of $\gtrsim$$150\,M_\odot$. It is located
in 30 Doradus at a projected distance of 29\,pc from the central
cluster R136. 
Its apparent isolation led to two intriguing hypotheses: 
either it formed in relative isolation 
or it was ejected dynamically from the cluster. 
We investigate the kinematics of
VFTS682 as obtained by \emph{Gaia} and multi-epoch \emph{Hubble Space Telescope} astrometry. We derive a projected velocity relative to
the cluster of $38\pm17\kms$ ($1\sigma$ confidence interval). Although
the error bars are substantial, two independent measures
suggest that VFTS682 is a runaway
ejected from the central cluster. This hypothesis is further supported by a variety of
circumstantial clues. The central cluster is known to harbor 
other stars more massive than $150\,M_\odot$ of similar spectral
type and recent astrometric studies on VFTS16 (and possibly VFTS72)
provide direct evidence that the cluster can eject some of its most
massive members, in agreement with theoretical predictions.
% . Recent analysis of the astrometry for the very massive O star VFTS16 (and possibly VFTS72), provides
% direct evidence that R136 can eject some of its
% most massive members, consistent with the predictions of numerical
% simulations. 
If future data confirm the runaway nature, this would make VFTS682
the most massive runaway star known to date. 
\end{abstract}

@@

In this paper we present an analysis of the new kinematic
constraints for VFTS682 provided by \emph{Gaia} DR2 and constraints
from HST proper motions  by \citet{platais:18}.   We discuss the
implications of the hypothesis that VFTS682 is a slow runaway star
ejected from R136.
% which was not part of the sample of \citet{lennon:18} because of its  spectral type.

@@

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.02\pm0.10$ & (5) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.19\pm0.09$ &  (5) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.19\pm0.09$ &  (5) \\
                  &$v_\mathrm{2D, HST}$  &[$\kms$] & $45\pm21$ & (5)\\  
                      &$\theta_\mathrm{HST}$  &[degrees] &   $-30_{-51}^{+24}$   & (1,6)\\  
                  %        &$v_\mathrm{2D, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
      \hline
      \multicolumn{5}{l}{\emph{Weighted average relative proper motion
      for VFTS682}}\\
      \hline
      &$\delta\mu_\mathrm{RA, avg}$  &[$\masyr$] & $0.08\pm0.07$ & (6) \\
      &$\delta\mu_\mathrm{DEC, avg}$  &[$\masyr$] & $0.14\pm0.07$ &  (6) \\
       &$\delta\mu_\mathrm{avg}$  &[$\masyr$] & $0.16\pm0.07$ &  (6) \\
      &$v_\mathrm{2D, avg}$  &[$\kms$] & $38\pm17$ & (6)\\
      % &$\theta_\mathrm{avg}$  &[degrees] &  \todo{TBD} $_{-}^{+}$   & (5)\\  
      \hline
       \multicolumn{5}{l}{\emph{Expected proper motion if ejected from
      R136 at age zero}} \\
      \hline
      &$v_\mathrm{2D}$  &[$\kms$] & $29\pm 6$ & (3) \\  
      &$\theta$  &[degrees] &  $\sim0$  & \\ 
       \hline
    \end{tabular}
    % \tablecomments
    % {

    % Assuming a distance of 50\,kpc, the conversion from proper motion
    % to projected velocities is such that $1\masyr$ corresponds to $237\kms$.
    {\tiny The error on the RA and DEC positions, are of order
      $\sim$$10^{-2}\masyr$ in \emph{Gaia}
      DR2. $\rho(\mu_\mathrm{RA},\mu_\mathrm{DEC})$ is the
      correlation coefficient. The position angle
      $\theta$ is defined such that $\theta=0$ for radial motion
      away from R136. We neglect the error bars on
      $\langle\mu_\mathrm{RA}\rangle_\mathrm{R136}$ and $\langle\mu_\mathrm{DEC}\rangle_\mathrm{R136}$ to determine the uncertainty
      on $\theta_{Gaia}$.
      (1)~\cite{brown:18},
      (2)~\cite{henault-brunet:12},
      (3)~\cite{bestenlehner:11},
      (4)~\cite{lennon:18}, 
      (5)~\cite{platais:18} and
      (6)~{\color{blue}this study}.
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

@@

VFTS682 is not a bright X-ray point source. It was not detected in the
\emph{Chandra} survey of \cite{townsley:06}, and shows a few counts in
the deeper survey of \cite{townsley:14}.
% \cite{townsley:06} did not find
% a bright X-ray point source at the position of VFTS682 in their
% \emph{Chandra} survey of the region.
The X-ray luminosity of VFTS682
is significantly lower than known massive binaries in the region, which suggests the absence of
colliding winds. These would be expected in the presence of companions
even for extreme mass ratios, given the extremely large mass of
VFTS682.

This star is also relatively isolated in the (near)
infrared. The nearest bright (near-)infrared sources detected by
\emph{Spitzer} \citep{meixner:06} and resolved in the \emph{VISTA}
Magellanic Clouds Survey \citep{cioni:11} are located at a distance of
about 10 arcsecond, i.e. about 2.4 pc. \cite{walborn:13} speculate
that these nearby young stars may represent a case of star formation triggered by the the wind of VFTS682.

@@

Estimates of the radial velocity are complicated by the variable,
possibly inhomogeneous, optically thick wind  typical of emission line
stars. \citet{bestenlehner:11} estimate a mass loss rate of
$\log_{10}(\dot{M}/[M_\odot \ \mathrm{yr}^{-1}])=-4.1\pm0.2$, not accounting for the
possible effect of clumping.  We therefore caution
against over-interpreting the radial velocities estimates. % , but we will summarize the estimates that have been published. 
\citet{bestenlehner:11}  estimate a RV of
$300\pm10\kms$ using the  N{\footnotesize V} $\lambda4944$ line, which
is offset from the average radial velocity of the region of
$270\pm10\kms$. This was suggested as indicating a runaway nature, but
it is no proof of it. % , and more recent measurements based on X-shooter
% data give different estimates for the RV ({\color{blue} Rubio-D{\'
%     i}ez et al. in prep}).
\cite{bressert:12} note an offset compared between the RV of
the star and the nebular lines from the gas filaments in its
vicinity. This is consistent with the expectation if the star was not
formed in situ. We will refrain from using the RV measurements in this
work, and focus on the velocities on the plane of the sky.

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}\\
  \vspace*{-5pt}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  % \caption{\emph{Top panel}: distribution of OB-type and Wolf-Rayet stars in proper
  %   motion relative to R136 from \emph{Gaia} DR2. VFTS682 is not an outlier, but
  %   its relative proper motion matches the expected value (green
  %   vertical band) if it were indeed
  %   ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. %The top axis shows the conversion to $\kms$
  %   % assuming a distance of 50\,kpc.
  %   \emph{Bottom panel}:  distribution of
  %   angles between the relative proper motion direction and the radial
  %   direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
  %   the best value is in agreement with the hypothesis of dynamical
  %   ejection (green vertical band). The peak at $\theta\simeq\pi/2$ is due to stars
  %   belonging to NGC2060. In both
  %   panels, the dark blue histograms contain 317 
  %   stars with error bars smaller than $0.1\,\mathrm{mas \
  %     yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
  %   lighter blue histograms contain 36 stars with error bars smaller than $0.05\,\mathrm{mas \
  %   yr^{-1}}$. }
  \caption{Distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136 (top panel) and proper motion angle
    (bottom panel), from \emph{Gaia} DR2. Although VFTS682 is not an outlier, its relative
    proper motion matches the value expected for an early dynamical ejection. In both
    panels, the green vertical bands highlight these expectations, the dark blue histograms contain 317 
    stars with error bars smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with error bars smaller than $0.05\,\mathrm{mas \
    yr^{-1}}$. The peak at $\theta\simeq\pi/2$ in the bottom panel is due to stars
  belonging to NGC2060. }
  \label{fig:dist}
\end{figure}

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figures/fig2}  
  % \caption{The thick yellow arrows indicate proper motions relative to
  %   R136 (blue
  %   circle) from \emph{Gaia} DR2, multiplied by 0.4\,Myr. The extensions in the opposite
  %   direction are proportional to the apparent ages of the stars, and
  %   the thin lines illustrate the error cone on the potential
  %   origin. The green thick arrow
  %   indicates the HST proper motion of VFTS682, with the relative
  %   error cone and extensions proportional to its apparent
  %   age. The two blue circles indicate the regions of radii 0.01 and 0.05
  %   degrees around the core of R136.}
  \caption{The thick purple arrow shows proper motion
    relative to R136 for VFTS682 averaging the \emph{Gaia} DR2 and HST
  astrometry, multiplied by 0.4\,Myr. The extension in the opposite
  direction is proportional to the apparent age of the star, and the
  thin lines illustrate the error cone on the potential origin. The
  yellow arrows show the \emph{Gaia} DR2 results alone, and the green
  arrow the HST results alone. The two blue circles indicate the regions of radii 0.01 and 0.05
   degrees around the core of R136.}  
  \label{fig:main}
\end{figure}

\vspace*{-20pt}
\subsection{ \emph{Gaia} astrometry for VFTS682\label{data:gaia}}

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog %\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}
  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e., groups of observations separated from other groups by a gap of at
least four days, used in the astrometric solution is seventeen. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable.

Gaia provides absolute proper motions.  To determine the relative
proper motion with respect to R136, we follow  \citet{lennon:18} to
define the motion of the local frame of reference using the average
proper motion of nearby stars with reliable astrometric data.  They
selected bright ($G<17$) stars within 0.05 degrees of R136 and exclude
sources with proper motion error bars greater than $0.1\masyr$ in both
coordinates (see their  Sect.~2.1).  Using this definition of the
local frame, we compute the relative proper motion
$\delta\mu_\mathrm{RA}$ and $\delta\mu_\mathrm{DEC}$.   We also compute the total projected 2D velocity
$v_\mathrm{2D}$ and the angle $\theta$ between the direction of motion and the vector
connecting the center of R136 with the current position of VFTS682,
i.e. such that $\theta = 0$ is the angle corresponding to perfectly
radial motion away from R136. All kinematic quantities are provided in
\Tabref{tab:vfts682}.

\vspace*{-20pt}
\subsection{HST (WFC3/UVIS) astrometry for VFTS682}

@@

The black points in~\Figref{fig:dist} show the proper motion 
(left panel) and the projected flight direction (right panel) relative
to R136 of
VFTS682 from \emph{Gaia} DR2. $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion
(see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce close to radial ejections, i.e.~$\theta\simeq0$. The green
vertical bands highlight the expectations for these two quantities if VFTS682 were ejected from R136 early in its
lifetime. For comparison, we also show the relative
proper motion of VFTS16 and VFTS72 (gray points), and the distribution
in relative proper motion and flight direction for all the VFTS OB-type and Wolf-Rayet stars with
\emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
blue). Although the error bars are substantial and VFTS682 is not
an outlier compared to other OB-type and Wolf-Rayet stars, the agreement
suggests that the star is indeed a ``slow runaway'' as suggested by
\cite{bestenlehner:11}.

Subtracting
the mean motion of R136, we obtain relative proper motions (projected
velocities) of $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.19\pm 0.09\masyr (45 \pm 21/\kms )$. Both values
are consistent with each other, but also with no motion relative to R136
within $2\sigma$. The average (weighted with $1/\sigma^2$) of these two independent
measurements is $\delta \mu_\mathrm{avg}=0.16\pm0.07\masyr
(38\pm17\kms)$.

Figure~\ref{fig:main} shows the projected motion of VFTS682 relative to R136 on the
sky. We also show VFTS16 and VFTS72 %for completeness
\citep[see][]{lennon:18}. The thick yellow arrows are proportional to
the relative proper motion from \emph{Gaia} DR2, and the thick green
line illustrates the relative proper motion from HST. The error cone
on the direction of motion is illustrated by the corresponding
extension in the direction opposite to the motion, and we also show
the most likely origin of the stars accounting for their apparent age.
This figure illustrates that R136 is the most likely origin of these stars, although the large error bars
prevent a robust identification for VFTS682, and there is some tension
between the apparent age and the present day distance from the cluster
core for VFTS16 and VFTS72 \citep[][]{lennon:18}. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances along
the line of sight are
needed to contrain the flight direction in three dimensions.

>>>>

How very massive stars form is still an open question in
astrophysics.  VFTS682 is among the most massive stars known,
with an inferred initial mass of $\gtrsim$$150\,M_\odot$. It is located
in 30 Doradus at a projected distance of 29\,pc from the central
cluster R136. 
Its apparent isolation led to two hypotheses: 
either it formed in relative isolation 
or it was ejected dynamically from the cluster. 
We investigate the kinematics of
VFTS682 as obtained by \emph{Gaia} and multi-epoch \emph{Hubble Space Telescope} astrometry. We derive a projected velocity relative to
the cluster of $38\pm17\kms$ ($1\sigma$ confidence interval). Although
the error bars are substantial, two independent measures
suggest that VFTS682 is a runaway
ejected from the central cluster. This hypothesis is further supported by a variety of
circumstantial clues. The central cluster is known to harbor 
other stars more massive than $150\,M_\odot$ of similar spectral
type and recent astrometric studies on VFTS16 (and possibly VFTS72)
provide direct evidence that the cluster can eject some of its most
massive members, in agreement with theoretical predictions.
% . Recent analysis of the astrometry for the very massive O star VFTS16 (and possibly VFTS72), provides
% direct evidence that R136 can eject some of its
% most massive members, consistent with the predictions of numerical
% simulations. 
If future data confirm the runaway nature, this would make VFTS682
the most massive runaway star known to date. 
\end{abstract}

@@

In this paper we present an analysis of the new kinematic
constraints for VFTS682 provided by \emph{Gaia} DR2 and constraints
from HST proper motions  by \citet{platais:18}.   We discuss the
implications of the hypothesis that VFTS682 is a runaway star
ejected from R136.
% which was not part of the sample of \citet{lennon:18} because of its  spectral type.

@@

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.02\pm0.10$ & (5) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.19\pm0.09$ &  (5) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.19\pm0.09$ &  (5) \\
                  &$v_\mathrm{2D, HST}$  &[$\kms$] & $45\pm21$ & (5)\\  
                      &$\theta_\mathrm{HST}$  &[degrees] &   $-30_{-51}^{+24}$   & (1,6)\\  
                  %        &$v_\mathrm{2D, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
      \hline
      \multicolumn{5}{l}{\emph{Weighted average relative proper motion
      for VFTS682}}\\
      \hline
      &$\delta\mu_\mathrm{RA, avg}$  &[$\masyr$] & $0.08\pm0.07$ & (6) \\
      &$\delta\mu_\mathrm{DEC, avg}$  &[$\masyr$] & $0.14\pm0.07$ &  (6) \\
       &$\delta\mu_\mathrm{avg}$  &[$\masyr$] & $0.16\pm0.07$ &  (6) \\
      &$v_\mathrm{2D, avg}$  &[$\kms$] & $38\pm17$ & (6)\\
      % &$\theta_\mathrm{avg}$  &[degrees] &  \todo{TBD} $_{-}^{+}$   & (5)\\  
      \hline
       \multicolumn{5}{l}{\emph{Expected proper motion if ejected from
      R136 at age zero}} \\
      \hline
      &$v_\mathrm{2D}$  &[$\kms$] & $29\pm 6$ & (3) \\  
      &$\theta$  &[degrees] &  $\sim0$  & \\ 
       \hline
    \end{tabular}
    {\tiny The error on the RA and DEC positions, are of order
      $\sim$$10^{-2}\masyr$ in \emph{Gaia}
      DR2. Assuming a distance of 50\,kpc, the conversion from proper motion
    to projected velocities is such that $1\masyr$ corresponds to $237\kms$.
$\rho(\mu_\mathrm{RA},\mu_\mathrm{DEC})$ is the
      correlation coefficient. The position angle
      $\theta$ is defined such that $\theta=0$ for radial motion
      away from R136. We neglect the error bars on
      $\langle\mu_\mathrm{RA}\rangle_\mathrm{R136}$ and $\langle\mu_\mathrm{DEC}\rangle_\mathrm{R136}$ to determine the uncertainty
      on $\theta_{Gaia}$.
      (1)~\cite{brown:18},
      (2)~\cite{henault-brunet:12},
      (3)~\cite{bestenlehner:11},
      (4)~\cite{lennon:18}, 
      (5)~\cite{platais:18} and
      (6)~{\color{blue}this study}.
    }
  \end{center}
  \label{tab:vfts682}
\end{table}

@@

VFTS682 is not a bright X-ray point source. It was not detected in the
\emph{Chandra} survey of \cite{townsley:06}, and shows a few counts in
the deeper survey of \cite{townsley:14}.
% \cite{townsley:06} did not find
% a bright X-ray point source at the position of VFTS682 in their
% \emph{Chandra} survey of the region.
The X-ray luminosity of VFTS682
is significantly lower than known massive binaries in the region, which suggests the absence of
colliding winds. These would be expected in the presence of companions
even for extreme mass ratios, given the large mass of
VFTS682.

This star is also relatively isolated in the \mbox{(near-)infrared}. The nearest bright (near-)infrared sources detected by
\emph{Spitzer} \citep{meixner:06} and resolved in the \emph{VISTA}
Magellanic Clouds Survey \citep{cioni:11} are located at a distance of
about 10 arcsecond, i.e. about 2.4 pc. \cite{walborn:13} speculate
that these nearby young stars may represent a case of star formation triggered by the the wind of VFTS682.

@@

Estimates of the radial velocity are complicated by the variable,
possibly inhomogeneous, optically thick wind  typical of emission line
stars.  We therefore caution
against over-interpreting the radial velocities estimates. \citet{bestenlehner:11} estimate a mass loss rate of
$\log_{10}(\dot{M}/[M_\odot \ \mathrm{yr}^{-1}])=-4.1\pm0.2$, not accounting for the
possible effect of clumping.  % , but we will summarize the estimates that have been published. 
They estimate a RV of
$300\pm10\kms$ using the  N{\footnotesize V} $\lambda4944$ line, which
is offset from the average radial velocity of the region of
$270\pm10\kms$. This was suggested as indicating a runaway nature, but
it is no proof of it. % , and more recent measurements based on X-shooter
% data give different estimates for the RV ({\color{blue} Rubio-D{\'
%     i}ez et al. in prep}).
\cite{bressert:12} note an offset compared between the RV of
the star and the nebular lines from the gas filaments in its
vicinity. This is consistent with the expectation if the star was not
formed in situ. We will refrain from using the RV measurements in this
work, and focus on the velocities on the plane of the sky.

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}\\
  \vspace*{-5pt}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  % \caption{\emph{Top panel}: distribution of OB-type and Wolf-Rayet stars in proper
  %   motion relative to R136 from \emph{Gaia} DR2. VFTS682 is not an outlier, but
  %   its relative proper motion matches the expected value (green
  %   vertical band) if it were indeed
  %   ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. %The top axis shows the conversion to $\kms$
  %   % assuming a distance of 50\,kpc.
  %   \emph{Bottom panel}:  distribution of
  %   angles between the relative proper motion direction and the radial
  %   direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
  %   the best value is in agreement with the hypothesis of dynamical
  %   ejection (green vertical band). The peak at $\theta\simeq\pi/2$ is due to stars
  %   belonging to NGC2060. In both
  %   panels, the dark blue histograms contain 317 
  %   stars with error bars smaller than $0.1\,\mathrm{mas \
  %     yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
  %   lighter blue histograms contain 36 stars with error bars smaller than $0.05\,\mathrm{mas \
  %   yr^{-1}}$. }
  \caption{Distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136 (top panel) and proper motion position angle
    (bottom panel), from \emph{Gaia} DR2. Although VFTS682 is not an outlier, its relative
    proper motion matches the value expected for an early dynamical ejection. In both
    panels, the green vertical bands highlight these expectations, the dark blue histograms contain 317 
    stars with error bars smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with error bars smaller than $0.05\,\mathrm{mas \
    yr^{-1}}$. The peak at $\theta\simeq\pi/2$ in the bottom panel is due to stars
  belonging to NGC2060. }
  \label{fig:dist}
\end{figure}

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figures/fig2}  
  % \caption{The thick yellow arrows indicate proper motions relative to
  %   R136 (blue
  %   circle) from \emph{Gaia} DR2, multiplied by 0.4\,Myr. The extensions in the opposite
  %   direction are proportional to the apparent ages of the stars, and
  %   the thin lines illustrate the error cone on the potential
  %   origin. The green thick arrow
  %   indicates the HST proper motion of VFTS682, with the relative
  %   error cone and extensions proportional to its apparent
  %   age. The two blue circles indicate the regions of radii 0.01 and 0.05
  %   degrees around the core of R136.}
  \caption{The thick purple arrow shows the proper motion
    relative to R136 for VFTS682 from averaging the \emph{Gaia} DR2 and HST
  astrometry, multiplied by 0.4\,Myr. The extension in the opposite
  direction is proportional to the apparent age of the star, and the
  thin lines illustrate the error cone on the potential origin. The
  yellow arrows show the \emph{Gaia} DR2 results alone, and the green
  arrow the HST results alone. The two blue circles indicate the regions of radii 0.01 and 0.05
   degrees around the core of R136.}  
  \label{fig:main}
\end{figure}

\vspace*{-15pt}
\subsection{ \emph{Gaia} astrometry for VFTS682\label{data:gaia}}

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog %\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}
  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e., groups of observations separated from each other by at
least four days, used in the astrometric solution is seventeen. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable.

Gaia provides absolute proper motions.  To determine the
proper motion relative to R136, we follow  \citet{lennon:18} to
define the motion of the local frame of reference using the average
proper motion of nearby stars with reliable astrometric data.  They
selected bright ($G<17$) stars within 0.05 degrees of R136 and exclude
sources with proper motion error bars greater than $0.1\masyr$ in both
coordinates (see their  Sect.~2.1).  Using this definition of the
local frame, we compute the relative proper motion
$\delta\mu_\mathrm{RA}$ and $\delta\mu_\mathrm{DEC}$.   We also compute the total projected 2D velocity
($v_\mathrm{2D}$) and the angle $\theta$ between the direction of motion and the vector
connecting the center of R136 with the current position of VFTS682.
% i.e. such that $\theta = 0$ is the angle corresponding to perfectly
% radial motion away from R136.
All kinematic quantities are provided in
\Tabref{tab:vfts682}.

\vspace*{-15pt}
\subsection{HST (WFC3/UVIS) astrometry for VFTS682}

@@

The black points in~\Figref{fig:dist} show the proper motion 
(left panel) and the projected flight direction (right panel) relative
to R136 of
VFTS682 from \emph{Gaia} DR2. $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion
(see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce close to radial ejections, i.e.~$\theta\simeq0$. The green
vertical bands highlight the expectations for these two quantities if VFTS682 were ejected from R136 early in its
lifetime. For comparison, we also show the relative
proper motion of VFTS16 and VFTS72 (gray points), and the distribution
in relative proper motion and flight direction for all the VFTS OB-type and Wolf-Rayet stars with
\emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
blue). Although the error bars are substantial and VFTS682 is not
an outlier compared to other OB-type and Wolf-Rayet stars, the agreement
suggests that the star is indeed a runaway as suggested by
\cite{bestenlehner:11}.

Subtracting
the mean motion of R136, we obtain relative proper motions (projected
velocities) of $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.19\pm 0.09\masyr (45 \pm 21\kms )$. Both values
are consistent with each other, but also with no motion relative to R136
within $2\sigma$. The average (weighted with $1/\sigma^2$) of these two independent
measurements is $\delta \mu_\mathrm{avg}=0.16\pm0.07\masyr
(38\pm17\kms)$.

Figure~\ref{fig:main} shows the motion of VFTS682 relative to R136
projected on the
sky. We also show VFTS16 and VFTS72 %for completeness
\citep[see][]{lennon:18}. The thick yellow arrows are proportional to
the relative proper motion from \emph{Gaia} DR2, the thick green
line illustrates the relative proper motion from HST, and the thick
purple line shows the averaged result. The error cone
on the direction of motion is illustrated by the corresponding
extension in the direction opposite to the motion, and we also show
the most likely origin of the stars accounting for their apparent age.
This figure illustrates that R136 is the most likely origin of these stars, although the large error bars
prevent a robust identification for VFTS682, and there is some tension
between the apparent age and the present day distance from the cluster
core for VFTS16 and VFTS72 \citep[][]{lennon:18}. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances along
the line of sight are
needed to contrain the flight direction in three dimensions.
